---
title: "Time Series Forecasting and Analysis Report"
author: "Forecasting Framework"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
params:
  data_path: "data/daily-total-female-births.csv"
  train_ratio: 0.6
  frequency: 7
  forecast_horizon: 154
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center', fig.width = 10, fig.height = 6)

# Source modules
source("R/data_loader.R")
source("R/models.R")
source("R/plots.R")

library(knitr)
library(tseries)
library(forecast)
library(ggplot2)
```

## Executive Summary
This document provides a comprehensive analysis and forecast of the time series dataset loaded from ``r params$data_path``. 

The analysis steps include:
1. Data cleaning and preparation.
2. Exploratory time series analysis (stationarity, decomposition, autocorrelation).
3. Model fitting (Seasonal Naive, ETS, Holt-Winters, and Auto-ARIMA).
4. Validation and error metrics evaluation against a test holdout set.
5. Statistical diagnostics of residuals.
6. Multi-step ahead forecasting.

---

## 1. Data Exploration & Quality Checks

```{r load_data}
# Ensure data is present
if (params$data_path == "data/daily-total-female-births.csv") {
  download_default_data(params$data_path)
}

raw_df <- load_and_clean_data(params$data_path)
y <- create_ts_object(raw_df, frequency = params$frequency)

# Basic statistics
summary_stats <- summary(y)
```

The time series contains `r length(y)` observations. Below is the statistical summary of the values:

```{r summary_table, echo=FALSE}
kable(t(as.matrix(summary_stats)), col.names = c("Min", "1st Qu.", "Median", "Mean", "3rd Qu.", "Max"), caption = "Time Series Descriptive Statistics")
```

### 1.1 Visualizing the Time Series
The raw observations plotted over time to check for trends, seasonality, or obvious outliers:

```{r plot_raw, echo=FALSE}
plot_time_series(y, title = "Original Time Series Observations")
```

### 1.2 Stationarity Test
We perform the Augmented Dickey-Fuller (ADF) test to evaluate if the series is stationary.
- **Null Hypothesis ($H_0$):** The series has a unit root (is non-stationary).
- **Alternative Hypothesis ($H_a$):** The series is stationary.

```{r adf_test}
adf_res <- adf.test(y)
```

The ADF statistic is **`r round(adf_res$statistic, 4)`** with a p-value of **`r round(adf_res$p.value, 4)`**.
`r if(adf_res$p.value < 0.05) { "Since the p-value is less than 0.05, we reject the null hypothesis and conclude that the series is **stationary**." } else { "Since the p-value is greater than 0.05, we fail to reject the null hypothesis, indicating the series is **non-stationary**." }`

### 1.3 Time Series Decomposition
Applying Seasonal and Trend decomposition using LOESS (STL) to separate the series into its underlying component parts: trend, weekly seasonal pattern, and residual noise.

```{r plot_decomp, echo=FALSE}
plot_decomposition(y)
```

---

## 2. Model Fitting and Evaluation

We split the dataset into training and test holdout sets using a split ratio of `r params$train_ratio * 100`%.
- **Training Set Size:** `r round(length(y) * params$train_ratio)` observations.
- **Test Set Size (Forecast Horizon $h$):** `r length(y) - round(length(y) * params$train_ratio)` observations.

```{r split_data}
split <- split_train_test(y, train_ratio = params$train_ratio)
train <- split$train
test <- split$test
h <- split$h
```

We fit four forecasting models to the training data:
1. **Seasonal Naive (SNAIVE):** Projects the last observed seasonal cycle forward.
2. **ETS (Error, Trend, Seasonal):** State space exponential smoothing model.
3. **Holt-Winters (HW):** Triple exponential smoothing.
4. **Auto-ARIMA (ARIMA):** Automated Box-Jenkins modeling.

```{r fit_models}
fc_snaive <- fit_snaive_model(train, h = h)
fc_ets <- fit_ets_model(train, h = h)
fc_hw <- fit_holtwinters_model(train, h = h, seasonal = "additive")
fc_arima <- fit_arima_model(train, h = h)
```

### 2.1 Model Comparison & Accuracy Metrics
We compute the prediction errors on the test dataset.

```{r accuracy_comp, echo=FALSE}
acc_snaive <- evaluate_model_accuracy(fc_snaive, test)
acc_ets <- evaluate_model_accuracy(fc_ets, test)
acc_hw <- evaluate_model_accuracy(fc_hw, test)
acc_arima <- evaluate_model_accuracy(fc_arima, test)

# Combine test metrics
metrics <- rbind(
  SNAIVE = acc_snaive["Test set", c("ME", "RMSE", "MAE", "MPE", "MAPE", "MASE")],
  ETS = acc_ets["Test set", c("ME", "RMSE", "MAE", "MPE", "MAPE", "MASE")],
  HoltWinters = acc_hw["Test set", c("ME", "RMSE", "MAE", "MPE", "MAPE", "MASE")],
  ARIMA = acc_arima["Test set", c("ME", "RMSE", "MAE", "MPE", "MAPE", "MASE")]
)

kable(metrics, digits = 3, caption = "Model Accuracy Comparison on Test Holdout Set")
```

### 2.2 Forecast Plots
The following sections plot the forecasts against the actual test values:

### 2.2.1 Seasonal Naive Forecast
```{r plot_snaive, echo=FALSE}
plot_forecast_vs_actual(fc_snaive, test, "Seasonal Naive")
```

### 2.2.2 ETS Forecast
```{r plot_ets, echo=FALSE}
plot_forecast_vs_actual(fc_ets, test, "ETS")
```

### 2.2.3 Holt-Winters Forecast
```{r plot_hw, echo=FALSE}
plot_forecast_vs_actual(fc_hw, test, "Holt-Winters Additive")
```

### 2.2.4 Auto-ARIMA Forecast
```{r plot_arima, echo=FALSE}
plot_forecast_vs_actual(fc_arima, test, "Auto-ARIMA")
```

---

## 3. Residual Diagnostics (Auto-ARIMA)
A good model should have residuals that resemble white noise (uncorrelated, zero mean, constant variance). We examine the residuals of the Auto-ARIMA model.

```{r residual_plots, echo=FALSE}
res_plots <- plot_residuals_diagnostics(fc_arima)
```

```{r show_res_line, echo=FALSE}
res_plots$line
```

```{r show_res_acf_hist, echo=FALSE, fig.show="hold", out.width="50%"}
res_plots$acf
res_plots$hist
```

### 3.1 Portmanteau Test (Ljung-Box)
We run a Ljung-Box test to check if the residuals have significant autocorrelations.
- **Null Hypothesis ($H_0$):** Residuals are independent (no autocorrelation, white noise).
- **Alternative Hypothesis ($H_a$):** Residuals exhibit autocorrelation.

```{r ljung_box}
lb_test <- Box.test(residuals(fc_arima$model), lag = params$frequency * 2, type = "Ljung-Box")
lb_test
```

The Ljung-Box test p-value is **`r round(lb_test$p.value, 4)`**.
`r if(lb_test$p.value > 0.05) { "Since the p-value is greater than 0.05, we fail to reject the null hypothesis, indicating that the residuals are **uncorrelated** and consistent with white noise." } else { "Since the p-value is less than 0.05, we reject the null hypothesis, indicating that there is **residual autocorrelation** and the model could be further refined." }`

---

## 4. Final Forecast
Based on the metrics above, the best-performing model can be selected. The following table displays the final forecast points and confidence bands for the upcoming `r params$forecast_horizon` time periods.

```{r final_forecast_table, echo=FALSE}
fc_final <- forecast(auto.arima(y), h = params$forecast_horizon)
final_df <- data.frame(
  Point_Forecast = as.numeric(fc_final$mean),
  Lo_80 = as.numeric(fc_final$lower[, 1]),
  Hi_80 = as.numeric(fc_final$upper[, 1]),
  Lo_95 = as.numeric(fc_final$lower[, 2]),
  Hi_95 = as.numeric(fc_final$upper[, 2])
)
kable(head(final_df, 15), digits = 2, caption = "First 15 Steps of Final Model Forecast Predictions")
```
